---
title: "Lab 09: Algorithmic Bias"
author: "Holland Sun"
date: "March 2026"
output: github_document
---

## Load Packages and Data  

First, let's load the necessary packages:  

```{r load-packages, message = FALSE}
library(tidyverse)
library(fairness)
compas <- read_csv("data/compas-scores-two-years.csv")
```
```{r}
identical(compas$decile_score...12, compas$decile_score...40)
if (identical(compas$decile_score...12, compas$decile_score...40)) {
  compas$decile_score <- compas$decile_score...12
  compas$decile_score...12 <- NULL
  compas$decile_score...40 <- NULL
}
identical(compas$priors_count...49, compas$priors_count...15)
if (identical(compas$priors_count...49, compas$priors_count...15)) {
  compas$priors_count <- compas$priors_count...15
  compas$priors_count...15 <- NULL
  compas$priors_count...49 <- NULL
}
```

Before We do the main exercise, the duplicate varaible should be merged.

## Exercises 1
The COMPAS dataset has ```r nrow(compas)``` rows and ```r ncol(compas)``` columns. Each row represents a defendant who was assessed by the COMPAS algorithm. The variables include ```r names(compas)```


## Exercises 2
```{r ex2-unique}
n_unique <- n_distinct(compas$id)
n_unique
nrow(compas)
```

There are `r n_distinct(compas$id)` unique defendants in the dataset, compared to `r nrow(compas)` rows. If these numbers differ, it may be because some defendants appear more than once due to multiple charges.

But they are same.

## Exercises 3

```{r ex3-score-dist}
ggplot(compas, aes(x = decile_score)) +
  geom_histogram(binwidth = 1, color = "black", fill = "steelblue") +
  labs(
    title = "Distribution of COMPAS Risk Scores",
    x = "Decile Score",
    y = "Count"
  ) +
  theme_minimal()
```


The distribution of COMPAS risk scores is right-skewed, with a large proportion of defendants receiving lower scores (1-3). This suggests the algorithm classifies the majority of defendants as relatively low risk, with fewer defendants receiving the highest scores.

## Exercises 4
```{r ex4-demographics}
# Refactor age_cat for proper ordering
compas <- compas %>%
  mutate(
    age_cat = factor(age_cat,
      levels = c("Less than 25", "25 - 45", "Greater than 45")
    )
  )

# race
ggplot(compas, aes(x = race, fill = sex)) +
  geom_bar(color = "black") +
  scale_fill_manual(values = c("steelblue", "lightsteelblue")) +
  labs(title = "Distribution of Defendants by Race and Sex",
       x = "Race", y = "Count") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))

# sex
ggplot(compas, aes(x = sex)) +
  geom_bar(fill = "steelblue", color = "black") +
  labs(title = "Distribution of Defendants by Sex", x = "Sex", y = "Count") +
  theme_minimal()

# age category
ggplot(compas, aes(x = age_cat)) +
  geom_bar(fill = "steelblue", color = "black") +
  labs(title = "Distribution of Defendants by Age Category", x = "Age Category", y = "Count") +
  theme_minimal()

# Combined visualization
ggplot(compas, aes(fill = age_cat, x = sex)) +
  geom_bar(position = "dodge") +
  facet_grid(~race) +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1)) +
  labs(x = "Sex", y = "Count", fill = "Age Category") +
    scale_fill_manual(values = c(
     "steelblue",
    "skyblue",
    "deepskyblue"
  ))
```

## Exercises 5
```{r ex5-score-recid}
recid_by_score <- compas %>%
  group_by(decile_score) %>%
  summarise(
    recid_rate = mean(two_year_recid, na.rm = TRUE),
    n = n()
  )

ggplot(recid_by_score, aes(x = decile_score, y = recid_rate)) +
  geom_col(fill = "steelblue", color = "black") +
  geom_text(aes(label = paste0(round(recid_rate * 100, 1), "%")), 
            vjust = -0.5, size = 3) +
  scale_x_continuous(breaks = 1:10) +
  scale_y_continuous(labels = scales::percent_format()) +
  labs(
    title = "Recidivism Rate by COMPAS Risk Score",
    x = "Decile Score",
    y = "Recidivism Rate"
  ) +
  theme_minimal()
```

Yes, higher risk scores do correspond to higher recidivism rates. Recidivism rate generally increases with the decile score. This suggests the COMPAS algorithm does have some predictive validity (though I suspect the relationship is not linear）.


## Exercises 6 

```{r ex6}
compas <- compas %>%
  mutate(
    compas_classification = case_when(
      decile_score >= 7 & two_year_recid == 1 ~ "TP",
      decile_score <= 4 & two_year_recid == 0 ~ "TN",
      decile_score >= 7 & two_year_recid == 0 ~ "FP",
      decile_score <= 4 & two_year_recid == 1 ~ "FN",
      TRUE ~ "Unclassified"
    )
  )

compas %>%
  count(compas_classification) %>%
  mutate(percentage = round(n / sum(n) * 100, 1))
```

## Exercises 7
```{r ex7-accuracy}
classified <- compas %>%
  filter(compas_classification != "Unclassified")

accuracy <- classified %>%
  summarise(
    total_classified = n(),
    correct = sum(compas_classification %in% c("TP", "TN")),
    accuracy = correct / total_classified
  )

accuracy
```

Among the classified cases (those with scores <= 4 or >= 7), the overall accuracy is approximately `r round(accuracy$accuracy * 100, 1)`%. This means the COMPAS algorithm makes the correct prediction for roughly half of the clear-cut cases. (but still having a huge false rate. so It only may server as a reference)

## Exercises 8
```{r ex8-score-by-race}
compas_br <- compas %>%
  filter(race %in% c("African-American", "Caucasian"))

ggplot(compas_br, aes(x = decile_score, fill = race)) +
  geom_histogram(binwidth = 1, position = "dodge", color = "black") +
  labs(
    title = "Distribution of Risk Scores by Race",
    x = "Decile Score",
    y = "Count",
    fill = "Race"
  ) +
    scale_fill_manual(values = c(
    "African-American" = "steelblue",
    "Caucasian" = "lightsteelblue"
  )) +
  scale_x_continuous(breaks = 1:10) +
  theme_minimal()
```

There is a clear difference in the distributions. Black defendants are more evenly spread across the risk score range and receive higher scores on average, while white defendants are heavily concentrated at lower scores (1-3). And they gradually decrease with the scores increasing.

### Exercise 9

```{r ex9-high-risk-rate}
compas_br %>%
  group_by(race) %>%
  summarise(
    total = n(),
    high_risk = sum(decile_score >= 7),
    percentage_high_risk = round(high_risk / total * 100, 1)
  )
```

More percentage of black defendants are classified as high risk compared to white defendants.(more than twice) 

But becasue they have different total amount. I guess we'd better do a chi-square test here:
```{r ex9-chi-test}
compas_br %>%
  mutate(high_risk = decile_score >= 7) %>%
  count(race, high_risk)
tab <- table(compas_br$race, compas_br$decile_score >= 7)
chisq.test(tab)
```
So now we may have pretty much confidence (as showed in p value) to say that there *may be* a *systmatic bias* here.

## Exercises 10 & 11
```{r ex1011-error-rates}

fpr_by_race <- compas_br %>%
  filter(two_year_recid == 0) %>%
  group_by(race) %>%
  summarise(
    total_non_recid = n(),
    false_positives = sum(decile_score >= 7),
    fpr = round(false_positives / total_non_recid * 100, 1)
  )
fpr_by_race


fnr_by_race <- compas_br %>%
  filter(two_year_recid == 1) %>%
  group_by(race) %>%
  summarise(
    total_recid = n(),
    false_negatives = sum(decile_score <= 4),
    fnr = round(false_negatives / total_recid * 100, 1)
  )
fnr_by_race
error_rates <- bind_rows(
  fpr_by_race %>% select(race, rate = fpr) %>% mutate(metric = "False Positive Rate (%)"),
  fnr_by_race %>% select(race, rate = fnr) %>% mutate(metric = "False Negative Rate (%)")
)

ggplot(error_rates, aes(x = race, y = rate, fill = race)) +
  geom_col(color = "black") +
  scale_fill_manual(values = c("steelblue", "lightsteelblue")) +
  geom_text(aes(label = paste0(rate, "%")), vjust = -0.5, size = 3.5) +
  facet_wrap(~metric) +
  labs(
    title = "Error Rate Disparities by Race",
    x = "Race",
    y = "Rate (%)",
    fill = "Race"
  ) +
  theme_minimal() +
  theme(legend.position = "none")
```

Black defendants who did not go on to reoffend were much more likely to be incorrectly classified as high risk (higher false positive rate). Conversely, white defendants who did reoffend were more likely to be incorrectly classified as low risk (higher false negative rate). 



## Exercises 12
```{r ex12-priors-race}
ggplot(compas_br, aes(x = decile_score, y = priors_count, color = race)) +
  geom_jitter(alpha = 0.1, width = 0.2, height = 0.2) +
  geom_smooth(method = "lm", se = FALSE) +
  labs(
    x = "Number of Prior Convictions",
    y = "COMPAS Decile Score",
    color = "Race"
  ) +
  scale_color_manual(values = c(
  "African-American" = "steelblue",
  "Caucasian" = "skyblue"
))+
  theme_minimal()
```

The fitting lines suggest that for a similar number of prior convictions, Black defendants tend to receive slightly higher risk scores than white defendants. This indicates the algorithm may be weighting prior convictions differently, or that other factors related to the race affect the model.

## Exercises 13
```{r ex13}
calibration <- compas_br %>%
  group_by(race, decile_score) %>%
  summarise(
    recid_rate = mean(two_year_recid, na.rm = TRUE),
    .groups = "drop"
  )

ggplot(calibration, aes(x = decile_score, y = recid_rate, color = race)) +
  geom_line(linewidth = 1) +
  geom_point(size = 2) +
  scale_x_continuous(breaks = 1:10) +
  scale_y_continuous(labels = scales::percent_format()) +
  labs(
    x = "Decile Score",
    y = "Actual Recidivism Rate",
    color = "Race"
  ) +
  scale_color_manual(values = c(
  "African-American" = "steelblue",
  "Caucasian" = "lightsteelblue"
))+
  theme_minimal()
```

From the perspective of Northpointe's calibration metrics, I think there is no obvious deviation between the two curves. If one group had consistently higher or lower recidivism rates across almost all score levels, their curves would show a clear difference. However, here, Black and White defendants have roughly similar recidivism rates. There is only a slight distinction around medium scores (3–5), but I believe this is normal. Therefore, from Northpointe's standpoint, there appears to be no issue.

## Exercises 14
I think there are several improvements we could consider. 

First, one idea would be to incorporate more features to reduce potential systematic biases in the data. Variables like race may partly reflect structural patterns in the justice system, so including additional relevant factors might help account for those influences rather than letting them distort the predictions.

Second, there is probably no permanent, one-time solution. Algorithms should be regularly audited by qualified professionals to evaluate how they perform across different demographic groups. In practice, There needs to be professionals who regularly audit algorithm performance across different demographic groups (I suspect such roles may already exist in reality; if not, this could represent another potential job opportunity for data scientists).

Third, it is important to weigh the trade-offs of different approaches. Methods and data are not inherently “right” or “wrong”(I think more may be talked in next exercise), and methods and data are neutral. What matters is how they are used and interpreted, which warrants deeper consideration. For instance, transforming a model's output into probabilities and conducting more careful analysis on cases with high uncertainty rather than automatic decisions and then imposing rules and constraints at the decision level which all could be more effective. However, this aspect likely involves many real-world issues beyond mathematical modeling. (I believe that why a competent data scientist also needs extensive knowledge in economics and sociology except mathematical)

## Exercises 15 & 16

I think different definitions of fairness can indeed conflict with each other mathematically. As we discussed earlier, a model that appears acceptable under one fairness criterion may show disparities under another (ProPublica vs. Northpointe). How to balance these tensions, however, involves broader practical considerations.

For example, we have to decide whether false positives (incorrectly labeling someone innocent as high risk) are more problematic than false negatives (failing to identify someoned angerous who may reoffend). These choices have real consequences, since different types of errors carry different social and ethical costs.

Risk assessment also relates to how public resources are allocated. With limited resources, I think models first are used to improve efficiency in deploying interventions or supervision. but with the deeper and board range we use them, we should spend more to enhance fairness, such as adding oversight mechanisms or constraints(as I mentioned in exerisece14), may themselves require additional resources. This raises the question of how to reconcile fairness objectives with resource limitations and competing policy priorities.

More broadly, these issues may reflect the implicit values or assumptions of a legal system or society. In different countries or cultural contexts, views about acceptable risks, priorities, and beneficiaries can vary, which further complicates the definition and implementation of algorithmic fairness. (I think this may involve much more complex political factors than I can think)


## Exercises 17
```{r ex17-priors-dist}
ggplot(compas_br, aes(x = priors_count, fill = race)) +
  geom_histogram(binwidth = 1, position = "dodge", color = "black") +
  labs(
    title = "Distribution of Prior Convictions by Race",
    x = "Number of Prior Convictions",
    y = "Count",
    fill = "Race"
  ) +
 scale_fill_manual(values = c(
  "African-American" = "steelblue",
  "Caucasian" = "lightsteelblue"
))+
  theme_minimal()
  

compas_br %>%
  group_by(race) %>%
  summarise(
    mean_priors = mean(priors_count, na.rm = TRUE),
    median_priors = median(priors_count, na.rm = TRUE)
  )
```

Black defendants in this dataset have a higher average number of prior convictions than white defendants. This difference partly explains the higher risk scores assigned to Black defendants. However, as I mentioned in the above exercise, it is important to note that prior convictions reflect not only individual behavior but also disparities in policing, prosecution, and sentencing practices.

### Exercise 18

```{r ex18}
ggplot(compas_br, aes(x = age, fill = race)) +
  geom_histogram(binwidth = 2, position = "dodge", color = "black", alpha = 0.7) +
  labs(title = "Age Distribution by Race", x = "Age", y = "Count", fill = "Race") +
   scale_fill_manual(values = c(
  "African-American" = "steelblue",
  "Caucasian" = "lightsteelblue"
))+
  theme_minimal()

ggplot(compas_br, aes(x = c_charge_degree, fill = race)) +
  geom_bar(position = "dodge", color = "black") +
  labs(title = "Charge Degree by Race", x = "Charge Degree (F=Felony, M=Misdemeanor)",
       y = "Count", fill = "Race") +
   scale_fill_manual(values = c(
  "African-American" = "steelblue",
  "Caucasian" = "lightsteelblue"
))+
  theme_minimal()
```

Black defendants in this sample tend to be younger and are more likely to face felony charges. Maybe younger age and felony charges are associated with higher risk scores in compas model.

### Exercise 19

```{r ex19-logistic-model}
# Create a logistic regression model
recid_model <- glm(
  two_year_recid ~ age + priors_count + c_charge_degree,
  data = compas,
  family = binomial()
)

# Add predicted probabilities to the dataset
compas <- compas %>%
  mutate(
    predicted_prob = predict(recid_model, type = "response"),
    our_high_risk = predicted_prob >= 0.5
  )
```

## Exercises 20
```{r ex20-our-model-fairness}
our_model_bw <- compas %>%
  filter(race %in% c("African-American", "Caucasian"))

# FPR for our model
our_fpr <- our_model_bw %>%
  filter(two_year_recid == 0) %>%
  group_by(race) %>%
  summarise(
    fpr = round(mean(our_high_risk) * 100, 1)
  )
our_fpr

# FNR for our model
our_fnr <- our_model_bw %>%
  filter(two_year_recid == 1) %>%
  group_by(race) %>%
  summarise(
    fnr = round(mean(!our_high_risk) * 100, 1)
  )
our_fnr
```

```{r ex20-compare-with-compas, fig.width=10, fig.height=5}
# COMPAS vs. Our Model
comparison <- bind_rows(
  fpr_by_race %>% select(race, rate = fpr) %>% mutate(metric = "FPR", model = "COMPAS"),
  fnr_by_race %>% select(race, rate = fnr) %>% mutate(metric = "FNR", model = "COMPAS"),
  our_fpr %>% select(race, rate = fpr) %>% mutate(metric = "FPR", model = "Our Model"),
  our_fnr %>% select(race, rate = fnr) %>% mutate(metric = "FNR", model = "Our Model")
)

ggplot(comparison, aes(x = race, y = rate, fill = model)) +
  geom_col(position = "dodge", color = "black") +
  geom_text(aes(label = paste0(rate, "%")),
            position = position_dodge(width = 0.9), vjust = -0.5, size = 3) +
  facet_wrap(~metric, labeller = labeller(metric = c(
    "FPR" = "False Positive Rate (%)",
    "FNR" = "False Negative Rate (%)"
  ))) +
  labs(
    title = "Error Rate Comparison: COMPAS vs. Our Logistic Regression Model",
    x = "Race", y = "Rate (%)", fill = "Model"
  ) +
  theme_minimal() +
  scale_fill_manual(values = c("COMPAS" = "coral", "Our Model" = "steelblue"))
```
Our model also seems to have a certain bias, but this is also reasonable. Just like the previous trick for you, these predictor variables (age, number of previous convictions, severity of charges) themselves are related to race.

## Exercises 21
```{r}
# Create a logistic regression model with race
recid_model_with_race <- glm(
  two_year_recid ~ age + priors_count + c_charge_degree + race,
  data = compas,
  family = binomial()
)

# Add predicted probabilities to the dataset
compas <- compas %>%
  mutate(
    predicted_prob_with_race = predict(recid_model_with_race, type = "response"),
    race_high_risk = predicted_prob_with_race >= 0.5
  )
```

## Exercises 22
```{r}
race_model_bw <- compas %>%
  filter(race %in% c("African-American", "Caucasian"))

# FPR for race model
race_fpr <- race_model_bw %>%
  filter(two_year_recid == 0) %>%
  group_by(race) %>%
  summarise(
    fpr = round(mean(race_high_risk) * 100, 1)
  )
race_fpr

# FNR for race model
race_fnr <- race_model_bw %>%
  filter(two_year_recid == 1) %>%
  group_by(race) %>%
  summarise(
    fnr = round(mean(!race_high_risk) * 100, 1)
  )
race_fnr

# RaceModel vs. Our Model
comparison <- bind_rows(
  our_fpr %>% select(race, rate = fpr) %>% mutate(metric = "FPR", model = "Our_model"),
  our_fnr %>% select(race, rate = fnr) %>% mutate(metric = "FNR", model = "Our_model"),
  race_fpr %>% select(race, rate = fpr) %>% mutate(metric = "FPR", model = "Race Model"),
  race_fnr %>% select(race, rate = fnr) %>% mutate(metric = "FNR", model = "Race Model")
)

ggplot(comparison, aes(x = race, y = rate, fill = model)) +
  geom_col(position = "dodge", color = "black") +
  geom_text(aes(label = paste0(rate, "%")),
            position = position_dodge(width = 0.9), vjust = -0.5, size = 3) +
  facet_wrap(~metric, labeller = labeller(metric = c(
    "FPR" = "False Positive Rate (%)",
    "FNR" = "False Negative Rate (%)"
  ))) +
  labs(
    title = "Error Rate Comparison: previous LRM vs. Race LRM",
    x = "Race", y = "Rate (%)", fill = "Model"
  ) +
  theme_minimal() +
  scale_fill_manual(values = c("Our_model" = "steelblue", "Race Model" = "deepskyblue"))
```

After adding race, the difference in error rates between groups did not disappear. Instead, there were changes in different directions in different indicators. It seems impossible to answer whether the changes are fair or unfair. On some levels, they are more fair and some are more unfair.

## Exercises 23

First, such tools should never be used as the sole basis for decisions about bail, sentencing, or parole. They should be one of many considerations and the judge must retain final discretion. Second, transparency is crucial: all the variables, training data and methodologies relied upon by any algorithm used for sentencing must be made public and subject to independent audit. Also the law should mandate regular fairness audits that evaluate differences in error rates across race, gender, and age groups. Fourth, defendants should have the right to know their risk score, understand how it is calculated, and have the right to challenge it. 